Ages of the Kepler Stars and Planets

### 2024-04-23 paper/ms.tex
update ms

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

%%%%%%%%%%%
% LITHIUM %
%%%%%%%%%%%
% number of KOIs that are not known FPs that have a HIRES spectrum
%\newcommand{\nkoiswithhires}{{1{,}500}}
% number of KOIs for which this work and Berger2018 overlap
%\newcommand{\nbergeroverlap}{{500}}

@@

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
%Ages of the Kepler Stars and Planets
\title{Ages of Stars and Planets Observed by Kepler}

@@

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded detailed
  pictures of how the rotation rates and lithium content of FGK stars
  evolve during the first few billion years of their lives.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is probably valid.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report new robust ages from both rotation and lithium for 14
  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
  super-Earths, three Earths and sub-Earths, and a Neptune-sized
  planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  The resulting age distribution bounds the relative balance of
  ``young'' and ``old'' planets detectable not only by Kepler, but
  also by future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting \footnote{See
\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Age distribution of stars (left) and known planet hosts
			(right) in the Kepler field inferred from rotation periods}.
		A histogram of samples from the age posteriors for all Kepler
		stars with reported rotation periods and temperatures is shown in
		light gray;
		the age calculation was actually valid for stars that meet the criteria
		discussed in Section~\ref{subsec:flags} (dark blue).
		\label{fig:hist_tgyro}
	}
\end{figure*}

@@

\begin{figure*}[!t]
	\begin{center}
%		\subfloat{
%			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
%		}
%		
%		\subfloat{
			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
%		}
	\end{center}
	\vspace{-0.5cm}

	\caption{
		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
		Planets highlighted in this plot have ages $t$<1\,Gyr at $2$$\sigma$.
%		
%		Two possible selection functions for what defines a young, sub-gigayear, planet
%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
%		Circles are previously reported measurements from the literature.
		Stars denote systems that are newly characterized in this work; circles
		are planets from the literature meeting the same requirement.
		Gray points are transiting planets from the NASA Exoplanet Archive older than
		1\,Gyr.
		The selection function matters:
		if we were to require $t$<1\,Gyr and $t/\sigma_{\rm t}>3$, we would loose $\approx$ten
		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
		\label{fig:rp_period_age_results}
	}
\end{figure*}

@@

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  Rows
  are sorted by the minimum of either the rotation or lithium-based
  age, and ``confirmed'' planets appear before``candidate'' planets.
  This version of the table is truncated to planets younger than
  one billion years.  The full machine-readable table contains ages
  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
  with MES$>$10.  The bit quality flags for the rotation-based ages,
  $Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main-sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  As an example, Kepler-1627 is flagged with bit 10 and bit 7.
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm grazing}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       bool  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
	%\hline
	\input{table_subgyr_candidate}
	\enddata
%	\tablecomments{This table includes X, Y, Z... The machine-readable version,
%		available online, includes additional columns for .... 
%	}
\end{deluxetable*}

>>>>

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

@@

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
%Ages of the Kepler Stars and Planets
\title{Ages of the Kepler Stars and Planets}

@@

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded detailed
  pictures of how the rotation rates and lithium content of FGK stars
  evolve during the first few billion years of their lives.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is probably valid.
  %
  The independent lithium ages show X,Y,Z, though they are subject to their own systematic issues.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report new robust ages from both rotation and lithium for 14
  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
  super-Earths, three Earths and sub-Earths, and a Neptune-sized
  planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  The resulting age distribution bounds the relative balance of
  ``young'' and ``old'' planets detectable not only by Kepler, but
  also by future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting.\footnote{See
\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Age distribution of stars (left) and planet hosts
			(right) in the Kepler field inferred from rotation periods}.
		The
		light gray histogram does not impose any requires on factors such as
		binarity, crowding, or the star's evolutionary state;
		the age calculation was more likely to be valid for stars that meet the criteria
		discussed in Section~\ref{subsec:flags} (black).
		\label{fig:hist_tgyro}
	}
\end{figure*}

@@

@@

\begin{figure*}[!t]
	\begin{center}
		%		\subfloat{
		%			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
		%		}
		%		
		%		\subfloat{
		\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
		%		}
	\end{center}
	\vspace{-0.5cm}

	\caption{
		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
		%		
		%		Two possible selection functions for what defines a young, sub-gigayear, planet
		%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
		%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
		%		Circles are previously reported measurements from the literature.
		Stars denote \nplyounggyrotwosigmanograzingnoruwe\ systems that are newly characterized in this work to have
		$t_{\rm gyro}$<1\,Gyr at $2$$\sigma$; circles
		are planets from the literature meeting the same age requirement, but with a heterogeneous
		set of age provenances.
		Gray points are transiting planets from the NASA Exoplanet Archive older than
		1\,Gyr.
		The selection function matters:
		selecting for precise ages ($t/\sigma_{\rm t}>3$) would hide $\approx$ten
		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
		For this plot, we require our Kepler stellar hosts to have no quality flags raised,
		and for the planets to similarly no have quality flags raised. 
		Weakening these constraints would expand the sample at the expense of reliability.
		\label{fig:rp_period_age_results}
	}
\end{figure*}

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates. 
  This version of the table is truncated to include the youngest fifty confirmed Kepler planets, and the twenty youngest candidates.
  The full machine-readable table contains ages
  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
  with MES$>$10.  
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       int  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
%	\rule
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{Rows
		are sorted by the minimum of either the rotation or lithium-based
		age, and ``confirmed'' planets appear before``candidate'' planets.
		The bit quality flags for the rotation-based ages,
		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
		Concisely summarized, they are:
		Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
		Bit 1: $\log g$$<$4.2?
		Bit 2: $M_{\rm G}$<3.9?
		Bit 3: In KEBC?
		Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
		Bit 5: Confused Kep-Gaia crossmatch?
		Bit 6: Gaia DR3 non-single star?
		Bit 7: RUWE$>$1.4?
		Bit 8: Crowded?
		Bit 9: Far from main-sequence?
		Bit 10: \citetalias{Santos_2021} CP/CB?
		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
		following meaning.
		Bit 0: Candidate not reliable?
		Bit 1: MES$<$10?
		Bit 2: Grazing?
	}
\end{deluxetable*}
